\section{Limitations}
\label{sct:limitations}

{\bf Type Annotations.} Using type annotations for annotating the compilation stage is not ideal. Type annotations are not fully
 integrated into the Scala language. Major drawbacks are that overloading resolution and implicit search are oblivious about annotations. For example, if two methods have the same signatures but different staging annotations the compiler will report an error. Implicit search will fail in two ways: \emph{i)} if two implicit functions with the same type are in scope but annotations differ the compiler will report an error about ambiguous implicits and \emph{ii)} if a method requires an implicit parameter with the \code{ct} annotation the compiler might provide an implicit argument without the annotation. These are not fundamental
 limitations and, in practice, stage polyvariance can be used to avoid difficulties with annotations.

{\bf Type Annotation Position.} Annotations in Scala can be used in many different positions and \tool supports only some of them. Annotation \code{ct} can not be used in following positions: \emph{i)} on classes, traits, and modules, \emph{ii) in the list of inherited classes and traits}, \emph{iii)} on the right hand side of the type variable definitions, and \emph{iv)} on all terms outside the method definitions (constructors, constructor arguments, etc.).

{\bf Access Modifiers.} Scala supports access modifiers of members. If methods that use \tool internally access \code{private} members of the class \tool will fail as all staged methods are inlined at the call site. Similar limitations exist with inlining functions in Scala. This problem could be circumvented inside Scala, however, the JVM will not allow this in the bytecode verification phase.

{\bf Code Explosion.} Annotation \code{ct} inlines all functions that are annotated and unrolls all loops on compile-time executed data structures. This can, for a staged function, lead to unnaceptable code explosion for some inputs while the code can behave regularly for other inputs. For example, calling \code{pow(0.5, 10000000)} on the function from \sct{sct:case-studies} will make unacceptably large code while \code{pow(0.5, 10)} will work as expected.

\subsection{Nominal Typing, Lower Bounds, and Higher-Kinded Types}
\label{sct:nominal-typing}
